Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-25 joint_inf.tex
Quibbles.

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
$$

>>>>

%% Useful packages
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{showlabels}

@@

\DeclareMathOperator*{\argmax}{argmax}

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \argmax_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

One tantalizing approach is to do away with the marginalization and directly estimate the maximum likelihood parameters from the fully-observed likelihood in \eqref{eq:full_likelihood}.
We can perform this by computing the profile likelihood
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\fullAlignment;\tau, t) = \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t),
\end{equation}
then estimating the topology and branch lengths via
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\fullAlignment;\tau, t)
\end{equation}
while using $\hat{\fullAncestralStates}$ maximizing \eqref{eq:profile_likelihood} as an estimate for $\fullAncestralStates$.
Since $\alphabet$ is a discrete set, there exists a maximum \eqref{eq:profile_likelihood}, and so \eqref{eq:profile_likelihood_topology_bl} will recover the joint ML values of the unknown parameters.
In the general case, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Neher2017}.

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
%EM Seems a little strange to give a new equation number to an equation that already exists with an assigned number.
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
$$
